---
title: "Kimberley Mound Springs invertebrate analyses"
author: "Adrian Pinder"
date and time: '`r Sys.time()`'
output: html_document
software: 'RStudio: Version 1.1.456 – © 2009-2018 RStudio, Inc. R version: `r getRversion()`'
editor_options: 
  chunk_output_type: console
---

Git repository https://github.com/AdrianMP62/KMS7analyses  

RStudio: Version 1.1.456 – © 2009-2018 RStudio, Inc. R version: `r getRversion()`  
Date and time: `r Sys.time()`

Uses the following datafiles:  
  
* ionic composition mgl.csv (2016-17 ionic composition data) with columns:   Year,CO3,Ca,Cl,HCO3,K,Mg,Na,SO4 (values as mg/L) and site codes (KMS01, KMS02 etc.) as row.headings  
* KMS17_invert_data.csv (2017 KMS invertebrate presence/absence data, with LOWESTID and LOWESTIDNC columns at front, then site codes)  
* ALL KMS-VBM data.csv (all KMS invertebrate presence/absence data - 1999-2017 with columns: "Class","Family","LOWESTID","LOWESTIDNC", then sample headings as year-two digit KMS code-sample type : P = Plankton, B = Benthic, PB = Plankton+Benthic, C = Core. Blanks already replaced with zeros  
* 

And undertakes the following analyses:  
  
* ordination of 2017 data  
* ionic composition ternary plots  
* cumulative richness by taxonomic group across all KMS data 1999 to 2017
* species accumulation for all KMS data 1999-201  
* Euler diagram for distribution of species across 3 sample types (plankton, benthic and core)


```{r setup}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(out.width='1.0\\linewidth')
set.seed(1234)
```

```{r}
options(scipen=999)

```

```{r, message=FALSE, results='hide'}
library(vegan)
library(ggplot2)
library(data.table)
library(simba)
library(Ternary)
library(ggplot2)
library(RColorBrewer)
library(iNEXT)
library(eulerr)
```


________________________________________________________________________________________  

##Ternary diagrams of ionic composition

```{r include=TRUE, message=FALSE}
#load data files
MgL.data <- read.csv("ionic composition mgl.csv", row.names=1)
MgL.data <- MgL.data[, 1:9]
```

```{r}
#create new data frame
Meq.data <- MgL.data
Meq.data$Na <- Meq.data$Na*0.0435
Meq.data$Ca <- Meq.data$Ca*0.0499
Meq.data$Mg <- Meq.data$Mg*0.0823
Meq.data$K <- Meq.data$K*0.0256
Meq.data$Cl <- Meq.data$Cl*0.0282
Meq.data$SO4 <- Meq.data$SO4*0.0208
Meq.data$HCO3 <- Meq.data$HCO3*0.0164
Meq.data$CO3 <- Meq.data$CO3*0.0333

```

```{r}
write.csv(Meq.data, "meq-data.csv")
```

```{r}
#calculate total anions and cations
Totals.cations <- rowSums(Meq.data[, colnames(Meq.data) %like% "Na|Ca|Mg|K"])
Totals.anions <- rowSums(Meq.data[, colnames(Meq.data) %like% "Cl|HCO3|CO3|SO4"])
```

```{r}
#calculate percentages
Meq.per <- data.frame(row.names = row.names(Meq.data))
Meq.per$Na <- Meq.data$Na/Totals.cations*100
Meq.per$Ca <- Meq.data$Ca/Totals.cations*100
Meq.per$Mg <- Meq.data$Mg/Totals.cations*100
Meq.per$K <- Meq.data$K/Totals.cations*100
Meq.per$Cl <- Meq.data$Cl/Totals.anions*100
Meq.per$SO4 <- Meq.data$SO4/Totals.anions*100
Meq.per$HCO3 <- Meq.data$HCO3/Totals.anions*100
Meq.per$CO3 <- Meq.data$CO3/Totals.anions*100
Meq.per$HC <- Meq.per$CO3 + Meq.per$HCO3
Meq.per$NaK <- Meq.per$Na + Meq.per$K
Meq.per$year <- MgL.data$Year

```

```{r}
Meq.per.16 <- Meq.per[Meq.per$year %like% "2016", ]
Meq.per.17 <- Meq.per[Meq.per$year %like% "2017", ]
```

```{r}
#create new datasets with just 3 anion and 3 cation variables for Ternary plots
Cations <- Meq.per[, colnames(Meq.per) %like% "NaK|Mg|Ca|year"]
row.names(Cations) <- c(1,8,10,11,12,13,14,15,16,4,19,20,21,22,23)
Anions <- Meq.per[, colnames(Meq.per) %like% "SO4|HC|Cl|year"]
Anions <- subset(Anions, select=-c(03)) #drop HCO3 from Anions but retain HC (=HCO3+CO3)
row.names(Anions) <- c(1,8,10,11,12,13,14,15,16,4,19,20,21,22,23)
```

```{r Ternary diagrams with Na+K and HCO3+CO3}
#create ternary diagrams
par(mfrow=c(1, 2), mar=rep(0.1, 4))
#first Ternary diagram
TernaryPlot(alab="Ca", blab="Mg", clab="NaK",
            point='up', lab.cex=1.2, grid.minor.lines = 0,
            grid.lty='solid', col=rgb(0.9, 0.9, 0.9), grid.col='white', 
            axis.col=rgb(0.6, 0.6, 0.6), ticks.col=rgb(0.6, 0.6, 0.6),
            padding=0.08)
AddToTernary(points, Cations[Cations$year %like% "2016", 1:3], pch=21, cex=symbol.size(Meq.per.16$Na/Meq.per.16$NaK, cex.max=5), col="red")
#AddToTernary(text, Anions, row.names(Cations), cex=0.7, font=2)
AddToTernary(points, Cations[Cations$year %like% "2017", 1:3], pch=21, cex=symbol.size(Meq.per.17$Na/Meq.per.17$NaK, cex.max=5), col="blue")
#AddToTernary(text, Anions, row.names(Cations), cex=0.7, font=2)
#AddToTernary(text, Cations, row.names(Cations), cex=0.7, font=2)
#second Ternary diagram
TernaryPlot(alab="Cl", blab="SO4", clab="HCO3+CO3",
            point='down', lab.cex=1.2, grid.minor.lines = 0,
            grid.lty='solid', col=rgb(0.9, 0.9, 0.9), grid.col='white', 
            axis.col=rgb(0.6, 0.6, 0.6), ticks.col=rgb(0.6, 0.6, 0.6),
            padding=0.08)
AddToTernary(points, Anions[Anions$year %like% "2016", 1:3], pch=21, cex=Meq.per.16$HCO3/Meq.per.16$HC*10, col="red")
#AddToTernary(text, Anions, row.names(Cations), cex=0.7, font=2)
AddToTernary(points, Anions[Anions$year %like% "2017", 1:3], pch=21, cex=Meq.per.17$HCO3/Meq.per.17$HC*10, col="blue")
#AddToTernary(text, Anions, row.names(Cations), cex=0.7, font=2)
```


________________________________________________________________________________________  
##Multivariate analyses of 2017 data only

<!-- use stringsAsfactors so the 1's are numbers -->
```{r upload data, results=FALSE}
data1 <- read.csv("KMS17_invert_data.csv", stringsAsFactors = FALSE, check.names = FALSE)
str(data1)
```

```{r strip data down to analysis matrix}
data2 <- data1[, -2]
row.names(data2) <-data2[, 1]
data2 <- data2[, -1]
species <- row.names(data2) #extract list of species
samples <- colnames(data2) #extract list of samples
data2 = as.matrix(as.data.frame(lapply(data2[, 1:26], as.numeric))) #convert data to numeric
row.names(data2) <- species
data2 <- t(data2)
row.names(data2) <- samples
```

```{r extract list of sample types}
sample.type <- substring(samples, 3,3)
sample.type <- gsub("P", "Plankton", sample.type)
sample.type <- gsub("B", "Benthic", sample.type)
sample.type <- gsub("C", "Core", sample.type)

```

```{r nMDS ordination in vegan, results=FALSE}
data2.mds <- metaMDS(data2, distance="bray")
data2.pts <- as.data.frame(data2.mds$points)
```

```{r}
data2.mds
```

```{r}
data2.mds.plot <- ggplot(data2.pts, aes(x=MDS1, y=MDS2)) + coord_fixed(ratio = 1)  + geom_point(aes(size=2, colour= sample.type)) + geom_text(aes(label = samples), vjust=0.5, hjust=1)
print(data2.mds.plot)
```

The above ordination plot shows differing composition of aquatic invertebrates by sample type, showing that the three sampling methods capture different components of the fauna. Site 14 (Potential Spring 1) is an open saline waterbody and has a very different fauna to the freshwater mound springs. The second core sample from site 11 (Attack Spring) was depauperate and that probbaly explains its separation from the other core samples. Otherwise there is no clear paterning amongst sites suggesting they are all variations on the same faunal theme.

The ordination below was run without singleton species but the resulting plot is almost identical to the one above (with singletons included) and with the same stress of about 0.17.

```{r, create new dataset by removing singleton species}
data2.no1s <- data2[, !(colSums(data2) == 1)] #requires creation of 'data2' from code above
```

```{r nMDS ordination in vegan, results=FALSE}
data2.no1s.mds <- metaMDS(data2.no1s, distance="bray")
data2.no1s.pts <- as.data.frame(data2.no1s.mds$points)
```

```{r plot new ordination}
data2.no1s.mds.plot <- ggplot(data2.no1s.pts, aes(x=MDS1, y=MDS2)) + coord_fixed(ratio = 1)  + geom_point(aes(size=2, colour= sample.type)) + geom_text(aes(label = samples), vjust=0.5, hjust=1)
print(data2.no1s.mds.plot)
```

```{r 3D nMDS ordination in vegan, results=FALSE, no singletons}
data2.no1s.3Dmds <- metaMDS(data2.no1s, k=3, distance="bray")
data2.no1s.3Dpts <- as.data.frame(data2.no1s.3Dmds$points)
```

```{r}
data2.No1s.3Dmds.plot <- ggplot(data2.no1s.3Dpts, aes(x=MDS2, y=MDS3)) + coord_fixed(ratio = 1)  + geom_point(aes(size=2, colour= sample.type)) + geom_text(aes(label = samples), vjust=0.5, hjust=1)
print(data2.No1s.3Dmds.plot)
```

________________________________________________________________________________________  

##Multivariate analysis of 2017 data combined across sample types (5 sites with plankton, benthic and core combined)

```{r upload data, results=FALSE}
data1 <- read.csv("KMS17_invert_data.csv", stringsAsFactors = FALSE, check.names = FALSE)
str(data1)
```

```{r strip data down to analysis matrix}
data2 <- data1[, -2]
row.names(data2) <-data2[, 1]
data2 <- data2[, -1]
species <- row.names(data2) #extract list of species
samples <- colnames(data2) #extract list of samples
data2 = as.matrix(as.data.frame(lapply(data2[, 1:26], as.numeric))) #convert data to numeric
row.names(data2) <- species
data2 <- t(data2)
row.names(data2) <- samples
```

```{r}
data3 <- data2[NULL, ]
data3[1, ] <- rowsum(data2, data2[grepl("01P|01C|01B", row.names(data2)), ])
```

```{r}
data3 <- data2[, ]
group <- data2[1:3, ]
data3[1, ] <- colSums(group)

```

```{r}
data2[27, ] <-colSums(data2))
x <- as.data.frame(colSums(data2))
x <- t(x)
data3[1, ] <- x
data3[1, ] <- x[2, ]
x <- sapply (data2)
```



```{r}
z <-mtcars
```



________________________________________________________________________________________  

##Calculate richness by major taxonomic group for all Kimberley Springs

```{r Read species matrix}
data.rich <- read.csv("ALL KMS-VBM data.csv", stringsAsFactors = FALSE, check.names = FALSE)
```

```{r create new empty matrix for richness values}
rich.sums <- data.rich[0, 4:NCOL(data.rich)]
rich.sums[1:13, 1] <- c("protozoans","other","rotifers","molluscs", "annelids","mites","cladocera","ostracoda","copepoda","amphipoda","shrimps","crabs","insects")
```

```{r use loop to calculte richness values per site for each of the twelve major taxon groups combined into one file}
row=0
for (i in c("^B","^I","^J","^K","^L","^M","^OG","^OH","^OJ","^OP","^OT","^OX","^Q"))
  {
  data.rich.temp <- data.rich[data.rich$LOWESTIDNC %like% i, ]
temp <- colSums(data.rich.temp[, 5:NCOL(data.rich)])
row=row+1
rich.sums[row, 2:NCOL(rich.sums)] <- temp
}
```

```{r calculate macrocrustacea and microcrustacea richness and delete individual crustacean orders. Note this assumes cladocerans, ostracods and copepods in rows 7 to 9 - CHECK IF MODIFYING THIS}
rich.sums[NROW(rich.sums)+1, 2:NCOL(rich.sums)] <- colSums(rich.sums[grepl("cladocera|ostracoda|copepoda", rich.sums$LOWESTIDNC), 2:NCOL(rich.sums)])
rich.sums[NROW(rich.sums), 1] <- "microcrustacea"
rich.sums[NROW(rich.sums)+1, 2:NCOL(rich.sums)] <- colSums(rich.sums[grepl("amphipoda|shrimps|crabs", rich.sums$LOWESTIDNC), 2:NCOL(rich.sums)])
rich.sums[NROW(rich.sums), 1] <- "macrocrustacea"
rich.sums <-  rich.sums[!grepl("cladocera|ostracoda|copepoda|amphipoda|shrimps|crabs",rich.sums$LOWESTIDNC),] 
```

```{r}
write.csv(rich.sums, "rich.sums.csv")
```

```{r add taxon names as row names and then convert to list format}
row.names(rich.sums) <- rich.sums[, 1]
rich.sums <- rich.sums[, -1]
rich.sums.list <- liste(t(rich.sums), x="taxon", y="sample", entry="richness")
```

###Compositional richness of  benthic samples collected from 1999-2001
```{r plot stacked column plot, 1990s}
ggplot(rich.sums.list[rich.sums.list$sample %like% "1999|2000|2001", ], aes(sample, richness, fill=factor(taxon,levels=c("protozoans","rotifers","other","molluscs","annelids","mites","microcrustacea","macrocrustacea","insects")))) + 
  scale_fill_brewer(palette="Set1") +  #use RColourBrewer package to set colours
  geom_bar(stat = "identity", colour="black") + #colour=black sets borders for the bars
  scale_y_continuous(limits=c(0, 150)) + #set y-limit to 100 for consistency
  guides(fill=guide_legend(title=NULL)) + #remove legend title
  theme_bw() + #remove background colour
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + #remove gridlines
  theme(axis.text.x=element_text(angle = -90, vjust = 0.5)) #angle x axis labels
```

###Compositional richness of combined plankton and benthic samples collected in 1993, 2003 and 2017
```{r plot stacked column plot, 1993 V-B samples, 2003 sample from Kachana and 2017 combined plankton and benthic}
ggplot(rich.sums.list[rich.sums.list$sample %like% "PB", ], aes(sample, richness, fill=factor(taxon,levels=c("protozoans","rotifers","other","molluscs","annelids","mites","microcrustacea","macrocrustacea","insects")))) + 
  scale_fill_brewer(palette="Set1") +  #use RColourBrewer package to set colours
  geom_bar(stat = "identity", colour="black") + #colour=black sets borders for the bars
  scale_y_continuous(limits=c(0, 150)) + #set y-limit to 150 for consistency
  guides(fill=guide_legend(title=NULL)) + #remove legend title
  theme_bw() + #remove background colour
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + #remove gridlines
  theme(axis.text.x=element_text(angle = -90, vjust = 0.5)) #angle x axis labels
```

###Compositional richness of plankton and benthic samples collected in 2017
```{r plot stacked column plot, 2017 benthic/plankton data}
ggplot(rich.sums.list[rich.sums.list$sample %like% "2017-B-|2017-P-|2017-P1|2017-P2", ], aes(sample, richness, fill=factor(taxon,levels=c("protozoans","rotifers","other","molluscs","annelids","mites","microcrustacea","macrocrustacea","insects")))) + 
  scale_fill_brewer(palette="Set1") +  #use RColourBrewer package to set colours
  geom_bar(stat = "identity", colour="black") + #colour=black sets borders for the bars
  scale_y_continuous(limits=c(0, 150)) + #set y-limit to 100 for consistency
  guides(fill=guide_legend(title=NULL)) + #remove legend title
  theme_bw() + #remove background colour
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + #remove gridlines
  theme(axis.text.x=element_text(angle = -90, vjust = 0.5)) #angle x axis labels
```

###Compositional richness of core samples collected in 2016 and 2017
```{r plot stacked column plot, 2016 and 2017 core samples}
ggplot(rich.sums.list[rich.sums.list$sample %like% "2016|2017-C", ], aes(sample, richness, fill=factor(taxon,levels=c("protozoans","rotifers","other","molluscs","annelids","mites","microcrustacea","macrocrustacea","insects")))) + 
  scale_fill_brewer(palette="Set1") +  #use RColourBrewer package to set colours
  geom_bar(stat = "identity", colour="black") + #colour=black sets borders for the bars
  scale_y_continuous(limits=c(0, 150)) + #set y-limit to 100 for consistency
  guides(fill=guide_legend(title=NULL)) + #remove legend title
  theme_bw() + #remove background colour
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + #remove gridlines
  theme(axis.text.x=element_text(angle = -90, vjust = 0.5)) #angle x axis labels
```


________________________________________________________________________________________  

##Species pool estimation

```{r load datafile which has species names in first column, and sites in columns}
invertspec <- read.csv("spec_accum.csv")
row.names(invertspec) <- invertspec[, 1]
invertspec <- invertspec[, -1]
invertspec <- t(invertspec)
```

```{r}
invertspec <- t(invertspec)
invert.spec.acc2 <- iNEXT(invertspec, q=0, endpoint=200, datatype="incidence_raw") #q=0 specifies species richness, endpoint = number of samples at which to stop extrapolation
plot(invert.spec.acc2)
```


________________________________________________________________________________________  

##Euler diagram showing species overlap between sample types

```{r draw euler plot with segment values defined by con. A,B,C are numbers of species that only occurred in benthic, plankton or core samples}
#based on 5 samples of each types collected at the same five springs
con <- c(A = 92, B = 63, C = 19, "A&B" = 57, "A&C"=14, "B&C" = 6, "A&B&C"=38)
plot(euler(con, shape = "ellipse"), edges = TRUE, quantities = TRUE, legend = list(labels = c("benthic", "plankton", "core"), cex=2))
```

###assess euler plot errors
```{r euler plot validation}
euler.calc <- euler(con, shape = "ellipse") #separate calculation of values for statitistical assessment of plot validity
error_plot(euler.calc, type = c("regionError", "residuals"), quantities = TRUE,
  pal = NULL) #plot errors
```


________________________________________________________________________________________  

##Species accumlation over time
```{r create dataframe}
spec.accum <- data.frame("Richness" = c(90, 165, 177, 179,	237,	303,	538))
spec.accum$Year <- as.Date(c('1993-10-17', '1999-9-10', '2000-10-01', '2001-07-01', '2003-09-06', '2016-05-27', '2017-08-04'))
```

```{r}
plot(spec.accum$Year, spec.accum$Richness, type="b", lwd=2, pch=10, xlab = "Year", ylab = "Cumulative Richness")
```





